---
title: "Monthly crime trends: a global STL-style decomposition"
author: "Andrew P. Wheeler"
format:
  pdf:
    documentclass: article
    geometry: margin=0.8in
  gfm: default
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(dplyr)
library(ggplot2)
library(readr)
source("src/plots.R")
global <- read_csv("src/data/model/global_stl.csv", show_col_types = FALSE)
metadata <- jsonlite::read_json("src/data/model/model_metadata.json", simplifyVector = TRUE)
global$date <- as.Date(global$date)
```

# What this project estimates

This is a descriptive decomposition of monthly crime counts for every city in
the RTCI national sample. Rates are annualized from monthly counts: a monthly
count `m` in a population `N` is shown as `12 * m / N * 100,000`. This puts
homicide on the familiar annual rate scale rather than below one.

The global model is an aggregated-binomial smooth decomposition. For crime
type `c`, city `i`, and month `t`, its global probability is:

`logit(p_global[c,t]) = alpha[c] + trend[c,t] + seasonal[c,month(t)]`.

The city fit then adds two empirical intercept terms to that global logit:
`city_effect[i]`, the population-weighted mean departure across crimes and
months, and `city_crime_effect[i,c]`, the population-weighted departure left
after the city effect. Thus:

`logit(p_city[i,c,t]) = logit(p_global[c,t]) + city_effect[i] + city_crime_effect[i,c]`.

The city-by-crime-by-month overdispersion term is the observed cell departure
from this fitted city baseline. It is centered at zero within every city and
crime type and retained at the complete time resolution. It is not a Pearson
residual. The global residual is centered at zero within crime type.

# Data coverage

The deliverable contains `r format(metadata$rows, big.mark = ",")` valid
city-month-crime observations from `r metadata$cities` cities. There is no
population threshold.

# Global trend

```{r}
plot_global_trend(global)
```

# Global seasonal component

```{r}
plot_global_seasonal(global)
```

# Global residual

```{r}
plot_global_residual(global)
```

All three figures use one readable panel per crime type. The complete
city-by-crime-by-month decomposition and the city-name crosswalk are available
in the app data; the app has crime selection, city detail, and a Leaflet map.
