---
title: "Monthly crime trends: a global STL-style decomposition"
author: "Andrew P. Wheeler"
format:
  pdf:
    documentclass: article
    geometry: margin=0.8in
  gfm: default
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(dplyr)
library(ggplot2)
library(readr)
source("src/plots.R")
global <- read_csv("src/data/model/global_stl.csv", show_col_types = FALSE)
metadata <- jsonlite::read_json("src/data/model/model_metadata.json", simplifyVector = TRUE)
global$date <- as.Date(global$date)
```

# What this project estimates

This is a descriptive decomposition of monthly crime counts for every city in
the RTCI national sample. Rates are annualized from monthly counts: a monthly
count Y in population N is displayed as 12 times Y divided by N, multiplied by
100,000. This puts homicide on the familiar annual rate scale rather than
below one.

The project uses one aggregated-binomial mixed model. For crime type (c),
city (i), and month (t),

$$
Y_{ict} \sim \operatorname{Binomial}(N_i,p_{ict}),
$$

$$
\operatorname{logit}(p_{ict}) = \alpha_c + f_c(t) + s_c(m_t) + b_i,
\qquad b_i \sim N(0,\sigma_b^2).
$$

Here alpha is the crime-specific intercept, f is a smooth crime-specific time
trend, s is a cyclic month-of-year effect, and b is the city random intercept.
The global prediction excludes b; the city prediction includes it.

For the requested city-by-crime-by-month overdispersion term, the observed
probability is stabilized as

$$
\tilde p_{ict}=\frac{Y_{ict}+0.5}{N_i+1},
\qquad
o_{ict}=\operatorname{logit}(\tilde p_{ict})-
\operatorname{logit}(p_{ict})-\bar o_{ic},
$$

where the barred o term is the population-weighted mean within city i and
crime type c. Therefore the overdispersion term is centered at zero for
every city and crime type. It is not a Pearson residual. The global residual is
constructed analogously from the global prediction and centered within crime
type.

# Data coverage

The deliverable contains `r format(metadata$rows, big.mark = ",")` valid
city-month-crime observations from `r metadata$cities` cities. There is no
population threshold.

# Global trend

```{r}
plot_global_trend(global)
```

# Global seasonal component

```{r}
plot_global_seasonal(global)
```

# Global residual

```{r}
plot_global_residual(global)
```

# Reading the trend and residual together

The trend is a smooth long-run level, while the residual is a short-run
departure from that level. Thus a declining murder trend can coexist with
positive recent residuals: recent observations can be above the currently
declining baseline even while the baseline itself remains below earlier years.
