---
title: "Monthly crime trends: global patterns and city departures"
author: "Andrew P. Wheeler"
bibliography: references.bib
link-citations: true
format:
  pdf:
    documentclass: article
    geometry: margin=0.8in
    header-includes:
      - \usepackage{float}
  gfm:
    fig-dpi: 192
    filters:
      - src/github-math.lua
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(dplyr)
library(ggplot2)
library(readr)
source("src/ensure_outputs_current.R")
rtci_ensure_outputs_current()
source("src/plots.R")
figure_path <- if (knitr::is_latex_output()) {
  "paper_files/figure-pdf/"
} else {
  "output/markdown/images/"
}
knitr::opts_chunk$set(fig_crop = FALSE, fig.path = figure_path)
global <- read_csv("src/data/model/global_stl.csv", show_col_types = FALSE)
decomposition <- read_csv("src/data/model/decomposition.csv", show_col_types = FALSE)
philly_burglary_source <- read_csv(
  "src/data/raw/rtci_crime_trends.csv",
  col_select = c(id, year, month, population, burglary_total),
  show_col_types = FALSE
) |>
  filter(id == "PAPEP0000", year == 2020) |>
  mutate(
    date = as.Date(sprintf("%04d-%02d-01", year, month)),
    annualized_rate = 12 * burglary_total / population * 100000
  )
philly_burglary_counts <- setNames(
  philly_burglary_source$burglary_total, philly_burglary_source$month
)
philly_burglary_rates <- setNames(
  philly_burglary_source$annualized_rate, philly_burglary_source$month
)
metadata <- jsonlite::read_json("src/data/model/model_metadata.json", simplifyVector = TRUE)
source_provenance <- jsonlite::read_json(
  "src/data/raw/source_metadata.json", simplifyVector = TRUE
)
city_components <- read_csv(
  "src/data/model/city_component_examples.csv", show_col_types = FALSE
)
seasonal_deviations <- read_csv(
  "src/data/model/city_seasonal_deviations.csv", show_col_types = FALSE
)
trend_deviations <- read_csv(
  "src/data/model/city_trend_deviations.csv", show_col_types = FALSE
)
latest_residual_se <- read_csv(
  "src/data/model/latest_residual_se.csv", show_col_types = FALSE
)
global$date <- as.Date(global$date)
decomposition$date <- as.Date(decomposition$date)
analysis_rows <- nrow(decomposition)
analysis_cities <- n_distinct(decomposition$city_id)
analysis_offenses <- n_distinct(decomposition$crime_type)
analysis_start_date <- min(decomposition$date, na.rm = TRUE)
analysis_end_date <- max(decomposition$date, na.rm = TRUE)
analysis_start_label <- format(analysis_start_date, "%B %Y")
analysis_end_label <- format(analysis_end_date, "%B %Y")
analysis_period_label <- paste(analysis_start_label, "through", analysis_end_label)
offense_end_dates <- decomposition |>
  group_by(crime_type) |>
  summarise(end_date = max(date, na.rm = TRUE), .groups = "drop")
latest_offense_text <- if (n_distinct(offense_end_dates$end_date) == 1L) {
  paste(
    "All", analysis_offenses, "offenses currently end in",
    format(offense_end_dates$end_date[[1]], "%B %Y")
  )
} else {
  paste(
    "The latest available offense months range from",
    format(min(offense_end_dates$end_date), "%B %Y"), "through",
    format(max(offense_end_dates$end_date), "%B %Y")
  )
}
source_downloaded_at <- as.POSIXct(
  source_provenance$downloaded_at_utc,
  format = "%Y-%m-%d %H:%M:%S",
  tz = "UTC"
)
if (is.na(source_downloaded_at)) {
  stop("The source download timestamp is missing or invalid.")
}
source_downloaded_label <- sub(
  " 0", " ", format(source_downloaded_at, "%B %d, %Y at %H:%M UTC"),
  fixed = TRUE
)
if (!identical(as.integer(metadata$rows), analysis_rows) ||
    !identical(as.integer(metadata$cities), analysis_cities)) {
  stop("Model metadata sample counts do not match the decomposition.")
}
if (!"global_residual_rate" %in% names(global) && "remainder_rate" %in% names(global)) {
  global$global_residual_rate <- global$remainder_rate
}
example_city <- "PAPEP0000"
example_crime <- "robbery"
example_data <- decomposition |>
  filter(city_id == example_city, crime_type == example_crime)
example_label <- unique(example_data$city_label)[1]
example_mean_departure <- mean(example_data$city_minus_global_logit, na.rm = TRUE)
example_min_departure <- example_data |>
  filter(is.finite(city_minus_global_logit)) |>
  slice_min(city_minus_global_logit, n = 1, with_ties = FALSE)
example_max_departure <- example_data |>
  filter(is.finite(city_minus_global_logit)) |>
  slice_max(city_minus_global_logit, n = 1, with_ties = FALSE)
example_peak_residual <- example_data |>
  filter(is.finite(overdispersion_logit)) |>
  slice_max(abs(overdispersion_logit), n = 1, with_ties = FALSE)
burglary_example_crime <- "burglary"
burglary_example_data <- decomposition |>
  filter(city_id == example_city, crime_type == burglary_example_crime)
burglary_peak_residual <- burglary_example_data |>
  filter(is.finite(overdispersion_logit)) |>
  slice_max(abs(overdispersion_logit), n = 1, with_ties = FALSE)
trend_changes <- global |>
  arrange(crime_type, date) |>
  group_by(crime_type) |>
  summarise(
    start_rate = first(trend_rate),
    end_rate = last(trend_rate),
    percent_change = 100 * (end_rate / start_rate - 1),
    .groups = "drop"
  )
trend_change <- function(crime) {
  trend_changes$percent_change[trend_changes$crime_type == crime]
}
crime_order <- c("murder", "rape", "robbery", "assault", "burglary", "theft", "motor")
trend_outliers <- trend_deviations |>
  group_by(crime_type) |>
  slice_max(robust_z, n = 2, with_ties = FALSE) |>
  ungroup() |>
  mutate(crime_type = factor(crime_type, levels = crime_order)) |>
  arrange(crime_type, desc(robust_z)) |>
  mutate(crime_type = as.character(crime_type))
seasonal_outliers <- seasonal_deviations |>
  group_by(crime_type) |>
  slice_max(robust_z, n = 2, with_ties = FALSE) |>
  ungroup() |>
  mutate(crime_type = factor(crime_type, levels = crime_order)) |>
  arrange(crime_type, desc(robust_z)) |>
  mutate(crime_type = as.character(crime_type))
residual_outliers <- decomposition |>
  filter(is.finite(overdispersion_logit)) |>
  group_by(crime_type, city_id, city_label) |>
  slice_max(abs(overdispersion_logit), n = 1, with_ties = FALSE) |>
  ungroup() |>
  group_by(crime_type) |>
  slice_max(abs(overdispersion_logit), n = 2, with_ties = FALSE) |>
  ungroup() |>
  mutate(crime_type = factor(crime_type, levels = crime_order)) |>
  arrange(crime_type, desc(abs(overdispersion_logit))) |>
  mutate(crime_type = as.character(crime_type))
latest_residual_outliers <- decomposition |>
  filter(is.finite(overdispersion_logit)) |>
  group_by(crime_type) |>
  filter(date == max(date, na.rm = TRUE)) |>
  slice_max(abs(overdispersion_logit), n = 2, with_ties = FALSE) |>
  ungroup() |>
  mutate(crime_type = factor(crime_type, levels = crime_order)) |>
  arrange(crime_type, desc(abs(overdispersion_logit))) |>
  mutate(crime_type = as.character(crime_type)) |>
  left_join(
    latest_residual_se |>
      mutate(date = as.Date(date)) |>
      select(crime_type, city_id, date, overdispersion_logit_se),
    by = c("crime_type", "city_id", "date")
  )
if (any(!is.finite(latest_residual_outliers$overdispersion_logit_se))) {
  stop("Latest residual standard errors are incomplete.")
}
outlier_table <- function(data, metric, digits = 3) {
  data |>
    group_by(crime_type) |>
    arrange(desc(robust_z), .by_group = TRUE) |>
    mutate(
      rank = row_number(),
      result = sprintf(
        paste0("%s (RMS %.", digits, "f; score %.2f)"),
        city_label, .data[[metric]], robust_z
      )
    ) |>
    ungroup() |>
    transmute(
      offense = crime_label(crime_type), rank = paste("City", rank), result
    ) |>
    tidyr::pivot_wider(names_from = rank, values_from = result) |>
    mutate(offense = factor(offense, levels = crime_label(crime_order))) |>
    arrange(offense) |>
    mutate(offense = as.character(offense))
}
trend_table <- outlier_table(trend_outliers, "trend_rms")
seasonal_table <- outlier_table(seasonal_outliers, "seasonal_rms")
residual_outlier_table <- function(data) {
  data |>
    group_by(crime_type) |>
    arrange(desc(abs(overdispersion_logit)), .by_group = TRUE) |>
    mutate(
      rank = paste("City", row_number()),
      result = sprintf(
        "%s (%s; e = %+.2f)", city_label, format(date, "%b %Y"),
        overdispersion_logit
      )
    ) |>
    ungroup() |>
    transmute(
      offense = crime_label(crime_type), rank, result
    ) |>
    tidyr::pivot_wider(names_from = rank, values_from = result) |>
    mutate(offense = factor(offense, levels = crime_label(crime_order))) |>
    arrange(offense) |>
    mutate(offense = as.character(offense))
}
residual_table <- residual_outlier_table(residual_outliers)
latest_residual_table <- latest_residual_outliers |>
  mutate(
    predicted_count = city_time_fitted_rate * population / (12 * 100000),
    offense = factor(crime_label(crime_type), levels = crime_label(crime_order))
  ) |>
  arrange(offense, desc(abs(overdispersion_logit))) |>
  transmute(
    offense = as.character(offense),
    city = city_label,
    residual = sprintf("%+.2f", overdispersion_logit),
    residual_se = sprintf("%.2f", overdispersion_logit_se),
    predicted_count = sprintf("%.1f", predicted_count),
    observed_count = format(
      round(count), big.mark = ",", trim = TRUE, nsmall = 0,
      scientific = FALSE
    ),
    predicted_rate = sprintf("%.1f", city_time_fitted_rate),
    observed_rate = sprintf("%.1f", observed_rate)
  )
```

# Introduction

Crime trends are often summarized with two numbers and a percent change. That
is convenient, but it makes normal volatility look important and hides whether
a change is temporary or sustained. Prior work has shown how time-series
graphs, prediction intervals, and fan charts give a more honest picture of
crime trends than binary before-and-after comparisons [@wheeler2016;
@wheeler2018; @yim2020]. Those tools answer whether a recent observation is
unusual. Here I address a related question: *what part of a city's change is
shared with other cities, and what part is local?*

There are at least three patterns worth separating. Crime can follow a gradual
trend shared across the country, a city can have a persistently different
trajectory or seasonal pattern, and an individual month can depart sharply
from both. Lumping those patterns into one residual term makes a short-lived
spike look like a new trend (or smooths away a local trend that is actually
different).

I use monthly reported crime from the Real-Time Crime Index (RTCI) [@rtci] and
fit a hierarchical binomial model separately for `r analysis_offenses`
offenses. The model has
a smooth trend and seasonal pattern shared across the sample, but it also gives
every city its own partially pooled trend and season. Two additional terms
separate monthly movement shared by all cities from the remaining city-month
variation.

The goal is descriptive. The model provides a practical way to compare the
national pattern, sustained city departures, and unusual months. It is not a
test of why crime changed, and it is not a forecasting model.

# Data and rate construction

I use `r format(analysis_rows, big.mark = ",")` city-month-offense observations
from `r analysis_cities` RTCI cities, covering `r analysis_offenses` offenses
from `r analysis_period_label`. I do not impose a population threshold.
For the figures, a monthly count $Y$ and population $N$ are displayed as
$12(Y/N)100{,}000$. This is an annualized rate per 100,000: it preserves the
observed monthly count but puts cities on a familiar scale.

The upstream files used for this build were downloaded on
`r source_downloaded_label`. Before rendering, the workflow checks the current
RTCI file revisions and atomically replaces only source files whose contents
changed. A content-signature change triggers the sequential model workflow;
otherwise the validated checkpoints are reused. Exact upstream revisions and
local checksums are stored with the source metadata and city crosswalk. This
keeps the paper and application reproducible while separating reported counts
from model-generated output. Data preparation uses R and the tidyverse
[@R2026; @tidyverse2019].

# Model

The practical goal is to distinguish level, trend, season, common monthly
movement, and local monthly noise. I fit a separate aggregated-binomial
generalized linear mixed model for each offense. This is equivalent to a
stacked model with no coefficients shared across offenses, but it avoids
building all `r analysis_offenses` random-effect systems at the same time. For city $i$ in
month $t$,

$$
Y_{it} \sim \mathrm{Binomial}(N_{it},p_{it}),
$$

$$
\mathrm{logit}(p_{it}) =
\alpha + \mathbf B(t)^\mathsf{T}\boldsymbol\beta
+ \mathbf F(m_t)^\mathsf{T}\boldsymbol\gamma
+ b_i + \mathbf B(t)^\mathsf{T}\mathbf u_i
+ \mathbf F(m_t)^\mathsf{T}\mathbf v_i + a_t + e_{it}.
$$

The global intercept is $\alpha$. The vector $\mathbf B(t)$ contains five
natural cubic spline basis functions evaluated at time $t$, with boundary and
interior knots fixed from the complete analysis period. Its fixed coefficients
$\boldsymbol\beta$ define the smooth global long-run trajectory. The vector
$\mathbf F(m_t)$ contains sine and cosine pairs for the first three annual
harmonics; $\boldsymbol\gamma$ therefore defines a smooth seasonal curve that
joins continuously from December to January. Natural cubic regression splines
provide a low-rank representation of gradual change, while paired Fourier terms
provide a direct cyclic representation of seasonality [@wood2017].

The city random intercept $b_i \sim N(0,\sigma_b^2)$ captures persistent
between-city differences in level. Each city also has its own trend-basis
coefficients $\mathbf u_i$ and seasonal-basis coefficients $\mathbf v_i$:

$$
\mathbf u_i \sim N(\mathbf 0,\sigma_u^2\mathbf I_5), \qquad
\mathbf v_i \sim N(\mathbf 0,\sigma_v^2\mathbf I_6).
$$

Thus the city trend is the global trend plus
$\mathbf B(t)^\mathsf{T}\mathbf u_i$, and the city seasonal curve is the
global seasonal curve plus $\mathbf F(m_t)^\mathsf{T}\mathbf v_i$. These are
the city-varying smooth terms: partial pooling shrinks weakly supported
departures toward the global curves instead of fitting an unrelated curve to
every city. This global-plus-deviation construction follows the logic of
factor-smooth comparisons described by @simpson2017 and hierarchical GAMs more
generally [@pedersen2019], implemented here through explicit low-rank random
coefficients.

The time-period intercept $a_t \sim N(0,\sigma_a^2)$ is shared by all
cities observed in month $t$. It captures nonsmooth national monthly departures
after the global trend and cyclic seasonal curve have been fitted. The final
term is an observation-level random effect,
$e_{it} \sim N(0,\sigma_e^2)$, unique to city $i$ in month $t$. It accommodates
extra-binomial variation without changing the binomial response distribution.
The distinction matters. The term $a_t$ moves every city in a given month,
whereas $e_{it}$ belongs to one city and one month.

In the fitted `glmmTMB` model, the corresponding computational specification is

```r
cbind(count, trials - count) ~ 1 +
  trend_b1 + ... + trend_b5 +
  season_s1 + season_c1 + ... + season_s3 + season_c3 +
  (1 | city_id) + (1 | time_period) +
  homdiag(0 + trend_b1 + ... + trend_b5 | city_trend_group) +
  homdiag(0 + season_s1 + ... + season_c3 | city_season_group) +
  (1 | cell_id)
```

Here `homdiag` assigns one estimated variance to all coefficients in a basis
block, sets their correlations to zero, and allows the coefficients to vary by
city. Despite their names, `city_trend_group` and `city_season_group` do not
combine cities into artificial groups. Each is a one-to-one copy of `city_id`
with the same `r analysis_cities` levels. The copies let `glmmTMB` represent the city intercept,
city trend coefficients, and city seasonal coefficients as three independent
covariance blocks; every city still receives its own five trend and six
seasonal coefficients. `homdiag` is a parsimonious ridge penalty: replacing it with `diag` would
estimate a separate variance for each basis coefficient, while an unstructured
block would also estimate their correlations. The homogeneous assumption is
computationally stable but depends on the scaling of the chosen basis. The
natural-spline knots and boundary knots are cached with each fitted
object, as are the three seasonal harmonics, so later predictions use exactly
the estimation bases. Models are estimated by restricted maximum likelihood
with `glmmTMB`, whose Template Model Builder backend uses automatic
differentiation and sparse random-effect calculations [@brooks2017;
@kristensen2016]. Each complete fitted model is saved and reload-verified before
the next offense begins.

Predictions are reported at four levels. The global curve contains the
intercept and fixed trend and seasonal basis terms. The city curve additionally
contains $b_i+\mathbf B(t)^\mathsf{T}\mathbf u_i+
\mathbf F(m_t)^\mathsf{T}\mathbf v_i$. The shared time-period remainder adds
$a_t$, and the final fitted value adds $e_{it}$. There is no continuity
correction or constructed "stabilized probability"; zeros are valid binomial
outcomes and all displayed components come directly from the fitted model.

## Computational implementation

I first implemented the model with `mgcv` factor smooths through `gamm4`: global
thin-plate and cyclic smooths were combined with city-specific `fs` trend and
seasonal smooths, while `lme4` represented the city, time-period, and
observation-level intercepts [@wood2011; @wood2017; @gamm4; @bates2015]. That
model worked on reduced sets of cities and years. The full offense data were a
different problem. Linux runs used more than 20 GB of memory before completing,
and Windows runs repeatedly terminated inside the sparse `Matrix`
factorization. A model that works only on a large rented machine is not a useful
default for this project.

The final implementation retains the structure that motivated the
factor-smooth model---a global intercept, global trend and season, city random
intercepts, city-varying trend and season, a common month effect, and an
observation-level effect---but changes its computational representation. The
explicit five-column natural-spline and six-column Fourier bases make the size
of each city's coefficient block fixed and transparent. Fitting those blocks
with `glmmTMB` reduced the full model to a sparse random-coefficient problem
that could be estimated and reload-verified one offense at a time. This is not
a general claim that `mgcv` factor smooths are infeasible.
It is a practical result for this combination of `r analysis_cities` cities,
`r format(analysis_rows, big.mark = ",")` city-month-offense observations, two
city-varying bases, and an
observation-level random effect.

# Global trends

@fig-global-trend shows the gradual change shared across cities after removing
seasonal, city-specific, and monthly variation. I use a separate vertical scale
for each offense; otherwise property crime would flatten the lower-rate murder
and rape series. Figures are produced with `ggplot2` [@wickham2016]. From
`r analysis_period_label`, the fitted burglary trend declined by
`r sprintf("%.1f", abs(trend_change("burglary")))`%, robbery by
`r sprintf("%.1f", abs(trend_change("robbery")))`%, murder by
`r sprintf("%.1f", abs(trend_change("murder")))`%, and theft by
`r sprintf("%.1f", abs(trend_change("theft")))`%. Assault was the exception:
its fitted global trend ended `r sprintf("%.1f", trend_change("assault"))`%
above its `r analysis_start_label` level. These endpoint comparisons summarize the curves;
they are not causal estimates. The paths between the endpoints are more useful
than the percentage alone.

```{r}
#| label: fig-global-trend
#| fig-width: 7
#| fig-height: 7.8
#| out-width: 100%
#| fig-pos: H
#| fig-cap: "Estimated global long-run trend for each offense, expressed as an annualized rate per 100,000. Each panel has its own vertical scale so that the shape of lower-rate offenses remains visible. The curves exclude city intercepts, city-specific smooth departures, and observation-level residuals."
plot_global_trend(global)
```

\clearpage

# Global seasonal patterns

@fig-global-seasonal answers a simple question: after accounting for the
long-run trend, which months tend to be higher or lower? Values above zero are
months above the fitted trend, and values below zero are months below it. The
effect is displayed on the annualized rate scale, so its magnitude depends on
the baseline rate for the offense as well as the logit-scale seasonal curve.

```{r}
#| label: fig-global-seasonal
#| fig-width: 7
#| fig-height: 7.8
#| out-width: 100%
#| fig-pos: H
#| fig-cap: "Estimated global cyclic seasonal departure for each offense on the annualized rate scale. Zero denotes the corresponding global trend; positive values indicate months above that trend and negative values indicate months below it. Panels use offense-specific vertical scales."
plot_global_seasonal(global)
```

\clearpage

# A city example: Philadelphia

## Robbery

Philadelphia robbery illustrates what each layer adds. The global curve is the
robbery pattern shared across the sample. The city curve adds Philadelphia's
level, trend, and seasonal departures. The observed series is much noisier
because it also includes shared monthly movement and the Philadelphia-specific
city-month residual.

```{r}
#| fig-width: 7
#| fig-height: 4.8
#| out-width: 100%
#| fig-pos: H
#| fig-cap: "Observed robbery rate and fitted global and Philadelphia curves."
plot_city_global_comparison(decomposition, example_city, example_crime)
```

Across the series, Philadelphia's fitted city curve differs from the global
curve by an average of `r sprintf("%.2f", example_mean_departure)` on the logit
scale. That average includes the city intercept. The departure ranges from
`r sprintf("%.2f", example_min_departure$city_minus_global_logit)` in
`r format(example_min_departure$date, "%B %Y")` to
`r sprintf("%.2f", example_max_departure$city_minus_global_logit)` in
`r format(example_max_departure$date, "%B %Y")`; movement within that range
reflects the city-specific trend and seasonal deviations. The largest fitted
monthly row residual occurs in `r format(example_peak_residual$date, "%B %Y")`.
It is not folded back into either smooth curve, which keeps an unusual month
from being mistaken for a persistent local trajectory. @fig-city-components
compares like components directly. The trend panel removes intercepts and
centers both curves, so it compares shape rather than level. The seasonal panel
superimposes Philadelphia's fitted seasonal pattern on the global pattern. The
residual panel compares the time-period effect shared by the full sample with
Philadelphia's total monthly residual, $a_t+e_{it}$. Shaded ribbons around the
Philadelphia trend and seasonal curves are approximate 95% pointwise
conditional intervals for the city-specific coefficient deviations.

```{r}
#| label: fig-city-components
#| fig-width: 7
#| fig-height: 6.5
#| out-width: 100%
#| fig-pos: H
#| fig-cap: "Philadelphia and full-sample robbery components on the logit scale: centered long-run trends with the city intercept removed (top), Fourier seasonal curves (middle), and the sample-wide time-period residual compared with Philadelphia's time-period plus city-month residual (bottom). Shading gives approximate 95% pointwise conditional intervals for Philadelphia's trend and seasonal deviations."
plot_city_component_comparison(city_components, example_city, example_crime)
```

\clearpage

## Burglary

Philadelphia burglary provides a useful contrast because several abrupt 2020
changes are visible in the reported series but should not be absorbed into the
city's long-run trend. In the RTCI source snapshot downloaded on
`r source_downloaded_label`, Philadelphia reported
`r format(philly_burglary_counts[["5"]], big.mark = ",")` burglaries in May,
`r format(philly_burglary_counts[["6"]], big.mark = ",")` in June, and
`r format(philly_burglary_counts[["10"]], big.mark = ",")` in October 2020.
Those counts correspond to annualized monthly rates of
`r sprintf("%.1f", philly_burglary_rates[["5"]])`,
`r sprintf("%.1f", philly_burglary_rates[["6"]])`, and
`r sprintf("%.1f", philly_burglary_rates[["10"]])` per 100,000,
respectively. These values are calculated directly from RTCI's reported count
and population fields, rather than from a model-generated artifact.

@fig-philly-burglary-rate shows that the partially pooled city curve remains
smooth through these spikes. This is intentional: the smooth trend and season
describe persistent structure, while isolated city-month departures belong to
the residual term. The largest fitted Philadelphia burglary row residual occurs
in `r format(burglary_peak_residual$date, "%B %Y")`.

```{r}
#| label: fig-philly-burglary-rate
#| fig-width: 7
#| fig-height: 4.8
#| out-width: 100%
#| fig-pos: H
#| fig-cap: "Philadelphia burglary rates: reported monthly observations, the global structural curve, and Philadelphia's fitted city curve. The pronounced May, June, and October 2020 observations come directly from the RTCI source counts; the smooth curves exclude shared time-period and city-month residual effects."
plot_city_global_comparison(
  decomposition, example_city, burglary_example_crime
)
```

\clearpage

@fig-philly-burglary-components separates the same series into trend, season,
and residual components. The centered trend panel removes Philadelphia's city
intercept; the seasonal panel compares like Fourier components; and the bottom
panel shows how the city-month term isolates the abrupt burglary departures
from the residual movement shared by the national sample.

```{r}
#| label: fig-philly-burglary-components
#| fig-width: 7
#| fig-height: 6.5
#| out-width: 100%
#| fig-pos: H
#| fig-cap: "Philadelphia and full-sample burglary components on the logit scale: centered long-run trends (top), Fourier seasonal curves (middle), and the sample-wide time-period residual compared with Philadelphia's time-period plus city-month residual (bottom). Approximate 95% pointwise conditional intervals are shown for Philadelphia's trend and seasonal deviations."
plot_city_component_comparison(
  city_components, example_city, burglary_example_crime
)
```

\clearpage

# Shared time period variation

@fig-global-residual shows the fitted common time-period effect $a_t$. I
transform it from the logit scale to a change in the annualized global rate.
This isolates monthly movement shared across the sample after accounting for
trend and season; it does not include the city-specific effect $e_{it}$. A
positive monthly effect can therefore occur during a declining trend whenever
that month is above its still-declining baseline.

```{r}
#| label: fig-global-residual
#| fig-width: 7
#| fig-height: 6.8
#| out-width: 100%
#| fig-pos: H
#| fig-cap: "Fitted shared time-period random effect for each offense, transformed to its change in the annualized global rate. Positive values are time-period departures above the global trend-plus-seasonal curve; negative values are departures below it. Panels use offense-specific vertical scales."
plot_global_residual(global)
```

The common time-period term separates abrupt movement from gradual change.
April 2020 is a prominent negative departure for rape, assault, and theft;
motor vehicle theft has a large positive shared departure in October 2023.
Their timing does not identify a cause.

\clearpage

# Cities with unusually different trends

A large city intercept only says that a city has a different average level. It
does not show that the city is moving in a different direction. To find unusual
trend shapes, I center each city's spline departure over time and summarize it
with its root-mean-square (RMS) magnitude. City-offense pairs
with fewer than 60 months were excluded. Log RMS was then standardized within
each offense using its median and median absolute deviation. Thus, every
offense contributes its two most unusual city trends, rather than allowing the
largest deviations from one offense to dominate the comparison. The ranking
excludes city intercepts, seasonal effects, shared time-period shocks, and
observation-level effects.

For each selected city, the conditional variances of its spline coefficients
were propagated through the centered trend basis. The shaded region is the
resulting estimate plus or minus 1.96 pointwise standard errors. Because
`glmmTMB` supplies diagonal conditional variances for these `homdiag` blocks,
the calculation omits posterior covariance among basis coefficients and
conditions on the fitted global curve and variance parameters. The ribbons are
therefore approximate pointwise uncertainty intervals, not simultaneous bands
or intervals for a new city's trajectory.

```{r}
#| tbl-cap: "Two cities with the largest standardized trend departures within each offense. RMS is calculated on the centered logit-scale spline departure; score is the within-offense robust standardized log RMS."
knitr::kable(trend_table, col.names = c("Offense", "First city", "Second city"))
```

@fig-trend-outliers-a and @fig-trend-outliers-b compare each selected city's
centered trend with the corresponding global trend. Centering removes both the
global intercept and the city random intercept, so separation between the lines
represents trajectory rather than average crime level.

```{r}
#| label: fig-trend-outliers-a
#| fig-width: 7
#| fig-height: 8.2
#| out-width: 100%
#| fig-pos: H
#| fig-cap: "The two most unusual city-specific trends within murder, rape, robbery, and assault. City and global spline trends are centered separately, removing intercept differences. Selection uses within-offense robust standardized RMS trend departure; seasonal and residual terms are excluded. Shading gives approximate 95% pointwise conditional intervals for each city's trend deviation."
plot_outlier_city_trends(
  city_components,
  filter(trend_outliers, crime_type %in% crime_order[1:4])
)
```

\clearpage

```{r}
#| label: fig-trend-outliers-b
#| fig-width: 7
#| fig-height: 8.2
#| out-width: 100%
#| fig-pos: H
#| fig-cap: "The two most unusual city-specific trends within burglary, theft, and motor vehicle theft. City and global spline trends are centered separately, removing intercept differences. Selection uses within-offense robust standardized RMS trend departure; seasonal and residual terms are excluded. Shading gives approximate 95% pointwise conditional intervals for each city's trend deviation."
plot_outlier_city_trends(
  city_components,
  filter(trend_outliers, crime_type %in% crime_order[5:7])
)
```

\clearpage

# Cities with different seasonal patterns

I rank seasonal departures separately from trend departures. For each
city-offense pair, I evaluate its six city-specific Fourier coefficients
over the 12 months and summarized by the RMS departure from the global
seasonal curve. Log RMS was standardized within offense by its median and
median absolute deviation. The two leading cities are retained separately for
every offense. This metric excludes city intercepts, long-run trend departures,
common monthly shocks, and observation-level effects.
Seasonal ribbons use the same diagonal conditional-variance propagation through
the centered Fourier basis and have the same pointwise, conditional
interpretation as the trend ribbons.

```{r}
#| tbl-cap: "Two cities with the largest standardized seasonal departures within each offense. RMS is calculated from the city-specific Fourier departure over the 12 months; score is the within-offense robust standardized log RMS."
knitr::kable(seasonal_table, col.names = c("Offense", "First city", "Second city"))
```

@fig-seasonal-outliers-a and @fig-seasonal-outliers-b superimpose every selected
city's fitted seasonal curve on the corresponding offense-wide curve.

```{r}
#| label: fig-seasonal-outliers-a
#| fig-width: 7
#| fig-height: 8.2
#| out-width: 100%
#| fig-pos: H
#| fig-cap: "The two most unusual city-specific seasonal patterns within murder, rape, robbery, and assault. Curves show the global and city-specific Fourier seasonal components over month of year on the logit scale; intercepts, trends, and residual terms are excluded. Shading gives approximate 95% pointwise conditional intervals for each city's seasonal deviation."
plot_seasonal_outlier_comparison(
  city_components,
  filter(seasonal_outliers, crime_type %in% crime_order[1:4])
)
```

\clearpage

```{r}
#| label: fig-seasonal-outliers-b
#| fig-width: 7
#| fig-height: 8.2
#| out-width: 100%
#| fig-pos: H
#| fig-cap: "The two most unusual city-specific seasonal patterns within burglary, theft, and motor vehicle theft. Curves show the global and city-specific Fourier seasonal components over month of year on the logit scale; intercepts, trends, and residual terms are excluded. Shading gives approximate 95% pointwise conditional intervals for each city's seasonal deviation."
plot_seasonal_outlier_comparison(
  city_components,
  filter(seasonal_outliers, crime_type %in% crime_order[5:7])
)
```

\clearpage

# Cities with unusually large monthly residuals

Trend and seasonal rankings describe persistent shapes. They are not designed
to identify one unusual month. For that task, I retain the month with the
largest absolute fitted observation-level effect $|e_{it}|$ within each city
and offense. I then select the two cities with the largest values separately
for each offense. Ranking one maximum per city ensures
that a single city cannot occupy both positions for an offense. Because
$e_{it}$ is on the common logit scale within each offense, the ordering does
not depend on city population or on converting the effect to an annualized
rate. It does, however, identify fitted residual extremes rather than raw-count
anomalies.

```{r}
#| tbl-cap: "Two cities with the largest absolute fitted city-month residual within each offense. The displayed month is the selected city's most extreme observation-level effect; e is the signed logit-scale residual."
knitr::kable(
  residual_table,
  col.names = c("Offense", "First city-month", "Second city-month")
)
```

@fig-residual-outliers-a and @fig-residual-outliers-b show the complete
city-month residual series for each selected city, with the ranked month marked
in red. These plots exclude the global trend, global season, city intercept,
city-specific trend and season, and shared time-period effect. Their vertical
scales are symmetric around zero within each panel so positive and negative
departures can be compared directly.

```{r}
#| label: fig-residual-outliers-a
#| fig-width: 7
#| fig-height: 8.2
#| out-width: 100%
#| fig-pos: H
#| fig-cap: "The two cities with the largest absolute fitted city-month residuals within murder, rape, robbery, and assault. Lines show the observation-level city-month effect over time; red points mark the month used for ranking. Panels are centered at zero and use city-specific symmetric vertical scales."
plot_residual_outlier_comparison(
  decomposition,
  filter(residual_outliers, crime_type %in% crime_order[1:4])
)
```

\clearpage

```{r}
#| label: fig-residual-outliers-b
#| fig-width: 7
#| fig-height: 8.2
#| out-width: 100%
#| fig-pos: H
#| fig-cap: "The two cities with the largest absolute fitted city-month residuals within burglary, theft, and motor vehicle theft. Lines show the observation-level city-month effect over time; red points mark the month used for ranking. Panels are centered at zero and use city-specific symmetric vertical scales."
plot_residual_outlier_comparison(
  decomposition,
  filter(residual_outliers, crime_type %in% crime_order[5:7])
)
```

\clearpage

# Most recent month residual outliers

The preceding ranking asks which cities had the most unusual fitted month at
any point in the series. For a current monitoring question, I instead restrict
the comparison to the latest available observation within each offense.
`r latest_offense_text`.
Within that single monthly cross-section, I rank cities by the absolute fitted
observation-level effect $|e_{it}|$ and retain the two largest values for each
offense. This identifies the clearest recent local departures after removing
the global and city-specific trends, seasonal terms, city intercept, and shared
time-period effect. The ranking is descriptive: it does not attach a separate
hypothesis test to each city or account for revisions to the most recent data.

For each selected city, @tbl-latest-residual-outliers compares the observed
outcome with the model-predicted mean before adding $e_{it}$. The prediction
therefore includes the city's intercept, trend, and seasonal terms plus the
shared time-period effect, but not the observation-level residual used to rank
the city. Counts are monthly. Rates are annualized from that month's count or
fitted mean and expressed per 100,000 residents.
The reported standard error for $e_{it}$ is the conditional standard deviation
of the fitted random effect, holding the estimated fixed effects and variance
parameters fixed. It describes uncertainty after partial pooling and is not a
standard error for the raw observed-minus-predicted difference.

```{r}
#| label: tbl-latest-residual-outliers
#| tbl-cap: "Observed outcomes and model-predicted means for the two cities with the largest absolute fitted city-month residual in the most recent available month for each offense. Pred. n and Obs. n are monthly counts; rate columns are annualized per 100,000 residents; e is the signed logit-scale residual."
knitr::kable(
  latest_residual_table,
  col.names = c(
    "Offense", "City", "$e$", "SE($e$)", "Pred. n", "Obs. n",
    "Pred. rate", "Obs. rate"
  ),
  align = c("l", "l", "r", "r", "r", "r", "r", "r"),
  escape = FALSE
)
```

@fig-latest-residual-outliers-a and @fig-latest-residual-outliers-b retain each
selected city's full residual history for context, while the red point marks
the latest observation used in the ranking. Consequently, earlier residuals
shown in these panels do not influence which cities were selected.

```{r}
#| label: fig-latest-residual-outliers-a
#| fig-width: 7
#| fig-height: 8.2
#| out-width: 100%
#| fig-pos: H
#| fig-cap: "The two largest absolute fitted city-month residuals in the most recent month within murder, rape, robbery, and assault. Lines provide each selected city's full residual history for context; red points mark the latest month used for ranking. Panels are centered at zero and use city-specific symmetric vertical scales."
plot_residual_outlier_comparison(
  decomposition,
  filter(latest_residual_outliers, crime_type %in% crime_order[1:4])
)
```

\clearpage

```{r}
#| label: fig-latest-residual-outliers-b
#| fig-width: 7
#| fig-height: 8.2
#| out-width: 100%
#| fig-pos: H
#| fig-cap: "The two largest absolute fitted city-month residuals in the most recent month within burglary, theft, and motor vehicle theft. Lines provide each selected city's full residual history for context; red points mark the latest month used for ranking. Panels are centered at zero and use city-specific symmetric vertical scales."
plot_residual_outlier_comparison(
  decomposition,
  filter(latest_residual_outliers, crime_type %in% crime_order[5:7])
)
```

# Discussion

The main lesson is to avoid forcing every change into one story. The global
smooth describes movement shared across the sample. A changing gap between a
city and that smooth indicates a different local trajectory. The residual term
then identifies months that neither smooth explains. Philadelphia burglary in
2020 is a clear example: June and October are important observations, but they
should not redefine the city's long-run curve.

This distinction extends the practical monitoring argument in my earlier work.
Percent changes can make ordinary low-count variation look dramatic, while a
single national series can hide substantial differences among cities
[@wheeler2016; @wheeler2018]. Fan charts address uncertainty around a forecast
[@yim2020]. The present model instead separates smooth common movement,
sustained city departures, and local monthly residuals. These are complementary
tasks, not competing definitions of a crime trend.

Partial pooling is what makes the city comparisons useful. If I estimated an
unconstrained curve for every city, the noisiest series would tend to look the
most distinctive. The shared variance parameters let well-supported city
patterns depart from the global curve while pulling unstable patterns toward
it [@pedersen2019; @simpson2017]. This is especially important when comparing
`r analysis_cities` cities with very different populations and crime counts.

A useful future extension would turn the residual screen into a formal
multiple-testing procedure. For each city-month effect, one could calculate a
standardized statistic from $e_{it}$ and its conditional standard error,
convert that statistic to a two-sided p-value, and then apply a false discovery
rate (FDR) correction across a prespecified family of cities and offenses. This
would address the fact that, when hundreds of cities are screened at once, some
apparently unusual residuals will occur by chance. Similar multiplicity issues
arise when the spatial point pattern test is used to identify local differences
across many areas; my prior work examined corrected local proportion tests, and
the Buffalo shooting analysis applied an FDR correction to the local spatial
comparisons [@wheeler2018sppt; @drake2023; @benjamini1995]. Because the proposed
tests would use shrunken random-effect estimates rather than independent raw
contrasts, their calibration should first be checked with simulation or a
parametric bootstrap.

The results are not causal effects or forecasts. Reported crime depends on
reporting practices, agency participation, revisions, and population
denominators. The binomial model gives counts and population a coherent link,
but it does not mean that residents have independent and identical risks. The
observation-level random effect accommodates extra variation; it does not tell
us why a particular month is unusual. I therefore view the estimates as a
monitoring and comparison tool. They show where to look next, not what caused
the pattern.

# Reproducibility and acknowledgments

The code, initial paper, and web application were created in OpenAI Codex using
Luna and Sol, with substantive review and direction from the author. Generated
metadata records the exact model formulas and settings, while source data,
geocoding results, and manual coordinate overrides are cached in the repository.

# References

\small
