---
title: "Monthly crime trends: a global STL-style decomposition"
author: "Andrew P. Wheeler"
format:
  pdf:
    documentclass: article
    geometry: margin=0.8in
  gfm: default
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(dplyr)
library(ggplot2)
library(readr)
source("src/plots.R")
global <- read_csv("src/data/model/global_stl.csv", show_col_types = FALSE)
metadata <- jsonlite::read_json("src/data/model/model_metadata.json", simplifyVector = TRUE)
global$date <- as.Date(global$date)
```

# What this project estimates

This is a descriptive decomposition of monthly crime counts for every city in
the RTCI national sample. Rates are annualized from monthly counts: a monthly
count `m` in a population `N` is shown as `12 * m / N * 100,000`. This puts
homicide on the familiar annual rate scale rather than below one.

The model estimates a crime-specific smooth long-run trend and a cyclic
month-of-year seasonal component for murder, rape, robbery, assault, burglary,
theft, and motor-vehicle theft. The global residual is the observed-versus-
global departure on the logit scale, centered at zero within crime type. The
city-level overdispersion term is retained separately in the complete app data
and is centered at zero within city and crime type.

# Data coverage

The deliverable contains `r format(metadata$rows, big.mark = ",")` valid
city-month-crime observations from `r metadata$cities` cities. There is no
population threshold.

# Global trend

```{r}
plot_global_trend(global)
```

# Global seasonal component

```{r}
plot_global_seasonal(global)
```

# Global residual

```{r}
plot_global_residual(global)
```

All three figures use one readable panel per crime type. The complete
city-by-crime-by-month decomposition and the city-name crosswalk are available
in the app data; the app has city and crime selectors plus a map view.
