---
title: "A spline decomposition of monthly crime trends"
author: "Andrew P. Wheeler"
format:
  pdf:
    documentclass: article
    geometry: margin=1in
  gfm: default
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
---

```{r}
knitr::opts_chunk$set(fig.path = "paper_files/figure-commonmark/")
library(dplyr)
library(ggplot2)
library(readr)
source("../src/plots.R")

global <- read_csv("../output/model/global_trends.csv", show_col_types = FALSE)
decomposition <- read_csv("../output/model/decomposition.csv", show_col_types = FALSE)
agency_summary <- read_csv("../output/model/agency_summary.csv", show_col_types = FALSE)
metadata <- jsonlite::read_json("../output/model/model_metadata.json", simplifyVector = TRUE)
global$date <- as.Date(global$date)
decomposition$date <- as.Date(decomposition$date)
```

# Overview

This paper develops a first-pass seasonal-trend decomposition for monthly
crime counts observed across agencies. The motivating idea is the aggregated
logit-binomial model used by Wheeler and Kovandzic [@wheeler2018monitoring],
extended so that the time structure is estimated with smooth functions. The
goal is descriptive: identify the common curve for each crime type, measure
agency departures from that curve, and flag unusually large monthly residuals.

The source snapshot comes from the Real-Time Crime Index repository
[@rtci]. The initial fit uses component crimes rather than aggregate violent
and property totals, which prevents the same component offenses from entering
the stacked model twice.

# Data and model

The analysis uses monthly agency-level counts for murder, rape, robbery,
assault, burglary, theft, and motor-vehicle theft. Rows are restricted to the
published `size == all` sample, the current population threshold, and valid
counts. The current fit uses a minimum population of
`r format(metadata$min_population, big.mark = ",")` and contains
`r format(metadata$rows, big.mark = ",")` stacked agency-month-crime
observations from `r metadata$agencies` agencies.

For agency \(i\), crime type \(c\), and month \(t\), let \(Y_{ict}\) be the
reported count and \(N_{it}\) the agency population. The response is

$$
Y_{ict} \sim \operatorname{Binomial}(N_{it}, p_{ict}), \qquad
\operatorname{logit}(p_{ict}) = \eta_{ict}.
$$

The linear predictor is

$$
\eta_{ict} = \alpha_c + f_c(\text{time}_{t}) + g_c(\text{year}_{t})
  + h_c(\text{month}_{t}) + b_i + b_{ic} + q_{ic}(\text{time}_{t}) + r_{ict}.
$$

Here \(f_c\) is the crime-specific global time curve, \(g_c\) is a smooth
year term, and \(h_c\) is a cyclic cubic spline for month. Agency and
agency-by-crime random effects provide partial pooling. The regularized
agency-by-crime random time slope \(q_{ic}\) estimates a persistent departure
from the global curve. The optional agency-month term \(r_{it}\), shared across
the stacked crime types, is the overdispersion term analogous to the `(1|Row)`
term in the motivating model. The published deliverable fit leaves this term
off for runtime on the full sample; focused runs can enable it with
`--overdispersion=true`.

The global baseline is obtained by predicting with agency and observation
terms excluded. Agency departures are the difference between the retained
agency prediction and this baseline on the logit scale. A monthly observation
is flagged when its absolute Pearson residual is at least 3; this threshold is
an exploratory screening rule, not a multiple-testing-adjusted hypothesis
test.

# Global trends

```{r}
plot_global_trends(global)
```

The thick curves are the population-weighted global spline predictions. The
thin curves show the corresponding observed rates. This view separates the
slowly evolving crime-specific baseline from month-to-month noise without
interpreting every local peak as a structural break.

# Agency departures

```{r}
plot_agency_deviations(decomposition, agency_summary, n = 14)
```

The figure displays agencies with the largest absolute fitted departures for
each crime type. A positive value means that the agency's fitted rate is above
the global crime-specific curve on the logit scale. The agency smooths are
regularized, so short isolated spikes are not automatically converted into a
new long-run trend.

# Residual screening

```{r}
plot_residuals(decomposition)
```

Residual screening is available at both the month and curve levels. The
browser app can filter a selected agency and crime type, change the date
window, and change the residual threshold. This makes it possible to inspect
whether a flagged month is isolated, part of a run, or aligned with a broader
departure in the agency curve.

# Limitations and next steps

The source data are reported crime counts and may contain missing reporting,
changes in agency coverage, and revisions. The current denominator is the
published population field; it is treated as the binomial number of trials for
this first implementation. The model is descriptive and does not identify
causal effects. In particular, the time spline, year spline, seasonality, and
agency smooths should be interpreted together rather than as independent
causal components.

Next steps are to add reporting-completeness diagnostics, compare alternative
denominators and quasi-binomial formulations, study sensitivity to population
thresholds, and automate a pinned monthly update workflow. The supplied
Wheeler paper and spline discussion by Simpson provide useful guidance for
prediction intervals and comparisons between smooths [@wheeler2018monitoring;
@simpson2017differencesplines].
