---
title: "Monthly crime trends: a global STL-style decomposition"
author: "Andrew P. Wheeler"
format:
  pdf:
    documentclass: article
    geometry: margin=1in
  gfm: default
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
---

```{r}
knitr::opts_chunk$set(fig.path = "paper_files/figure-commonmark/")
library(dplyr)
library(ggplot2)
library(readr)
source("../src/plots.R")
global <- read_csv("../output/model/global_stl.csv", show_col_types = FALSE)
decomposition <- read_csv("../output/model/decomposition.csv", show_col_types = FALSE)
metadata <- jsonlite::read_json("../output/model/model_metadata.json", simplifyVector = TRUE)
global$date <- as.Date(global$date)
decomposition$date <- as.Date(decomposition$date)
first_city <- decomposition$city_id[[1]]
first_crime <- as.character(decomposition$crime_type[[1]])
```

# What this project estimates

This is a descriptive decomposition of monthly crime counts for every city in
the RTCI national sample. Rates below are annualized from monthly counts, so a
monthly murder count of (m) in a population (N) is shown as
(12m/N \times 100{,}000). This is why the homicide values are on the familiar
annual rate scale rather than below one.

The model uses a single stacked aggregated-binomial response for murder, rape,
robbery, assault, burglary, theft, and motor-vehicle theft. It estimates a
crime-specific smooth long-run trend, a smooth year term, and a cyclic
month-of-year seasonal term. City and city-by-crime effects can be enabled for
focused runs. The national deliverable uses the stable global fit for every
city, then reports the requested city × crime × month overdispersion term as
the logit departure of the observed cell from that baseline.

The source is the pinned Real-Time Crime Index snapshot [@rtci], and the
aggregated-binomial motivation follows Wheeler and Kovandzic
[@wheeler2018monitoring].

# Data coverage

The deliverable contains `r format(metadata$rows, big.mark = ",")` valid
city-month-crime observations from `r metadata$cities` cities. It includes no
population threshold. The city-name and coordinate crosswalk comes from the
RTCI processed agency metadata.

# Global STL-style decomposition

```{r}
plot_global_stl(global)
```

The legend is explicit: grey is the observed population-weighted rate, blue is
the long-run smooth trend, and orange is trend plus the cyclic seasonal effect.

```{r}
plot_seasonal(global)
```

# City detail and overdispersion

```{r}
plot_city_detail(decomposition, first_city, first_crime)
plot_city_overdispersion(decomposition, first_city, first_crime)
```

The overdispersion term is not a Pearson residual and is not filtered by an
arbitrary threshold. It is retained at the city, crime, and month level so the
app can show the complete time series and let the analyst decide which
departures matter.

# Interpretation and next steps

The decomposition is descriptive, not causal. Reporting completeness,
population estimates, agency coverage, and revisions can all affect the
curves. The next useful extensions are uncertainty intervals for the global
and city terms, a formal shrinkage model for the city × crime × month effect,
and a sensitivity analysis for reporting changes.

